The Gender Pay Gap in the General Social Survey

### 2023-08-28 Paper/paper.tex
using log income everywhere

\usepackage[backend=biber, maxbibnames = 99, style = apa]{biblatex}
\setlength\bibitemsep{1.5\itemsep}
\addbibresource{R_packages.bib}
\bibliography{references.bib}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true, linktocpage = TRUE]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Alex Amaguaya, Lea Bergmann},
            pdftitle={The Gender Pay Gap in the General Social Survey},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

  Ape Monkey

@@

  Ape@Monkey.biz

  \nth{5}

@@

  Darth Vader

  543556

  vader@emperialenterprises.com

  \nth{3}

@@

The prediction of individuals' wages has been an issue in previous
literature. As wages are relevant to companies they want to know how
predict wages \autocite{Chakraborti}. Some papers make use of
classification techniques to predict wages \autocite{Chakraborti}. As we
have numeric values of real wages and not a categorical variable we do
not make use of classification techniques but prefer linear regression
techniques such as OLS and Lasso regressions as well as random forests.
We prefer these techniques as they handle categorical variables and
outliers well, which are features of our data \autocite{Cutler}. Wages
were already predicted using random forests by \autocite{Eichinger}.
They find that using random forests results in better predictions than
using linear models. Additionally, they show that gender barely effects
the wage predictions.

@@

The initial data set contains 11 variables, which are described in the
following: - \textbf{year}: survey year - \textbf{realrinc}:
respondent's base income (in constant 1986 USD) - \textbf{age}:
respondent's age in years - \textbf{occ10}: respondent's occupation code
- \textbf{occrecode}: recode of the occupation code into one of 11 main
categories - \textbf{prestg10}: respondent's occupational prestige score
- \textbf{childs}: number of children - \textbf{wrkstat}: work status of
the respondent (full-time, part-time, temporarily not working,
unemployed (laid off), etc.) - \textbf{gender}: respondent's gender -
\textbf{educcat}: respondent's degree level (Less Than High School, High
School, etc.) - \textbf{maritalcat}: respondent's marital status
(Married, Widowed, Divorced, etc.)

@@

Before modeling we prepared the data. The initial data set has 54,842
observations and the outcome variable \texttt{realrinc} has 38.55\% of
missing values. The imputation of the target variable was discarded to
avoid generating a bias in the modeling and therefore we decided to
eliminate these records from the original data set. After this, the data
set contains 33,702 observations. Furthermore, it was impossible to
deduce the occupation for some observations (\texttt{occ10\ =\ 9997})
and these are removed from the data set. At the end, the final data set
has 33,244 observations, and is used for the modeling process.

This data set still has some missing values for the regressors, e.g.,
number of children, age, marital status, degree level, and others. Using
the \texttt{mice} package we conduct the imputation using five
variables. We impute the features \emph{age, occupational prestige and
number of children} using the pmm (\texttt{predictive\ mean\ matching})
method. The polytomous logistic regression method is used for the factor
variables (\emph{education and marital status}). The imputation process
uses a number of multiple imputations of six. This means that the
imputation process creates 6 different data sets, and then the
estimation of the mean and the majority vote is applied to aggregate the
data sets into a single data set. The mean estimation is applied for
numeric variables and majority vote for categorical variables..

@@

After the imputation process we create interaction variables. First,
some interaction features are estimated using only the numerical
variables, and then we transform the numerical variables into
categorical variables by dividing them into different ranges (e.g.~age
between 18 and 30, age between 31 and 50, etc). These new categorical
variables and the initial categorical variables (\emph{marital status,
education, etc}) are transformed into dummy variables. And finally, we
estimate the interaction variables with the group of dummy variables
mentioned above. We validate the inclusion of the interaction of the
numeric variables using a linear model. The results show that the
p-values of the parameters for the interactions are less than 0.05.

\begin{figure}
\centering
\includegraphics{includes/reg_table.png}
\caption{table}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{includes/corr_matrix.png}
\caption{matrix\_corr}
\end{figure}

After the data preparation and feature engineering processes, we
construct a final data set with 33,244 observations and 421 features.
Due to the large number of interaction variables, we decided to reduce
them and use only the interaction between a group of variables
(\emph{occrecode, educcat and prestg10}). After reduction, the modeling
process is developed with 100 variables (numerical and interaction
variables). The following figure shows the overview of the group of
variables that we estimated and used in the models.

\begin{figure}
\centering
\includegraphics{includes/feature_process.png}
\caption{feature\_process}
\end{figure}

@@

We use five different regression techniques to predict the individuals'
wages using the final dataset. We fit every model on our training data
using the \texttt{train} function from the \texttt{caret} package. To
choose the best parameter combination we use cross-validation within all
train functions.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{formula }\OtherTok{\textless{}{-}} \FunctionTok{as.formula}\NormalTok{(}\FunctionTok{paste}\NormalTok{(}\StringTok{"realrinc \textasciitilde{}"}\NormalTok{, }\FunctionTok{paste}\NormalTok{(x\_cols\_dummys, }\AttributeTok{collapse =} \StringTok{" + "}\NormalTok{)))}
\NormalTok{mod\_full }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula,}
                  \AttributeTok{data =}\NormalTok{ train, }
                  \AttributeTok{method =} \StringTok{"lm"}\NormalTok{,  }
                  \AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Given some extreme outliers in the income distribution of our data
(which is typical for wage data) we follow the economic literature
\autocite{Ermini} and fit a logistic model to our data, using the log of
\texttt{realrinc} as our dependent variable. This results in a higher
RMSE than the linear model (25,909.82).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{formula2 }\OtherTok{\textless{}{-}} \FunctionTok{as.formula}\NormalTok{(}\FunctionTok{paste}\NormalTok{(}\StringTok{"log\_realrinc \textasciitilde{}"}\NormalTok{, }\FunctionTok{paste}\NormalTok{(x\_cols\_dummys, }\AttributeTok{collapse =} \StringTok{" + "}\NormalTok{)))}
\NormalTok{mod\_loginc }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula2,}
                    \AttributeTok{data =}\NormalTok{ train, }
                    \AttributeTok{method =} \StringTok{"lm"}\NormalTok{,  }
                    \AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\[
RSS + \lambda \sum^p _{j=1} \beta^2_j 
\]

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ridge }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula, }\AttributeTok{data =}\NormalTok{ train,}
  \AttributeTok{method =} \StringTok{"glmnet"}\NormalTok{, }\AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{),}
  \AttributeTok{tuneGrid =} \FunctionTok{expand.grid}\NormalTok{(}\AttributeTok{alpha =} \DecValTok{0}\NormalTok{, }\AttributeTok{lambda =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{500}\NormalTok{, }\DecValTok{1}\NormalTok{))) }
\end{Highlighting}
\end{Shaded}

Perform a ridge regression to predict real incomes yields an RMSE of
25,524.85.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lasso }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula, }\AttributeTok{data =}\NormalTok{ train, }
       \AttributeTok{method =} \StringTok{"glmnet"}\NormalTok{, }\AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{), }
       \AttributeTok{tuneGrid =} \FunctionTok{expand.grid}\NormalTok{(}\AttributeTok{alpha =} \DecValTok{1}\NormalTok{, }\AttributeTok{lambda =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{500}\NormalTok{,}\DecValTok{1}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

To perform the lasso regression in \texttt{R} we set \(\alpha\) equal to
one. The lasso regression yields a lower RMSE than the ridge regression
(25,504.58 vs.~25,524.85) and even a lower RMSE than the linear
regression.

In a final step we use elastic net regularization which is a linear
combination of ridge and lasso regression. The elastic net
regularization allows \(\alpha\) to vary between zero and one. This
results in an RMSE of 25,241.7.

@@

>>>>

\usepackage[backend=biber, maxbibnames = 99, style = apa]{biblatex}
\setlength\bibitemsep{1.5\itemsep}
\addbibresource{R_packages.bib}
\bibliography{ref.bib}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true, linktocpage = TRUE]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Alex Amaguaya, Lea Bergmann},
            pdftitle={The Gender Pay Gap in the General Social Survey},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

 Lea Bergmann

@@

lea.bergmann@rwi-essen.de
lea.bergmann.1j@stud.uni-due.de

PhD Economics

  \nth{1}

  Summer Term 2023

@@

@@

The prediction of individuals' wages has been an issue in previous
literature. As wages are relevant to companies they want to know how
predict wages \autocite{Chakraborti}. Some papers make use of
classification techniques to predict wages \autocite{Chakraborti}. As we
have numeric values of real wages and not a categorical variable we do
not make use of classification techniques but prefer linear regression
techniques such as OLS and Lasso regressions as well as random forests.
We prefer these techniques as they handle categorical variables and
outliers well, which are features of our data \autocite{Cutler}. Wages
were already predicted using random forests by \autocite{Eichinger}.
They find that using random forests results in better predictions than
using linear models. Additionally, they show that gender barely effects
the wage predictions. Likewise, the boosting methods used by
\autocite{Chakrabarty2018} in salary level classification problems
showed better performance compared to a traditional algorithm such as
SVM. Furthermore, \autocite{BonaccoltoToepfer1612} suggests the use of a
penalization approach for high dimensional problems in his research
about Gender wage inequality.

@@

The initial data set contains 11 variables, which are described in the
following:

\begin{itemize}
\item
  \textbf{year}: survey year
\item
  \textbf{realrinc}: respondent's base income (in constant 1986 USD)
\item
  \textbf{age}: respondent's age in years
\item
  \textbf{occ10}: respondent's occupation code
\item
  \textbf{occrecode}: recode of the occupation code into one of 11 main
  categories
\item
  \textbf{prestg10}: respondent's occupational prestige score
\item
  \textbf{childs}: number of children
\item
  \textbf{wrkstat}: work status of the respondent (full-time, part-time,
  temporarily not working, unemployed (laid off), etc.)
\item
  \textbf{gender}: respondent's gender
\item
  \textbf{educcat}: respondent's degree level (Less Than High School,
  High School, etc.)
\item
  \textbf{maritalcat}: respondent's marital status (Married, Widowed,
  Divorced, etc.)
\end{itemize}

@@

Before modeling we conduct a data preparation process. The initial data
set has 54,842 observations and the outcome variable \texttt{realrinc}
has 38.55\% of missing values. The imputation of the target variable is
discarded to avoid generating a bias in the modeling and therefore we
decided to eliminate these records from the original data set. After
this, the dataset contains 33,702 observations. Furthermore, it is
impossible to deduce the occupation for some observations
(\texttt{occ10\ =\ 9997}) and we remove these from the data set. At the
end, the final data set has 33,244 observations, and we use it for the
modeling process.

This data set still had some missing values for the regressors, e.g.,
number of children, age, marital status, degree level, and others. We
use the \texttt{mice} package with five variables for the imputation.
The features \emph{age, occupational prestige and number of children}
are imputed using the pmm (\texttt{predictive\ mean\ matching}) method.
We use the polytomous logistic regression method for the factor
variables (\emph{education and marital status}). The imputation process
generates 6 different data sets. We apply the estimation of the mean and
the majority vote to aggregate the data sets into a single data set. The
mean estimation is applied for the numeric variables and majority vote
for the categorical variables..

@@

After the imputation process, we create the interaction variables.
First, we estimate some interaction features using only the numerical
variables. Then, we transform the numerical variables into categorical
variables by dividing them into different ranges (e.g., age between 18
and 30, age between 31 and 50, etc.). We transform these new categorical
variables and the initial categorical variables (\emph{marital status,
education, etc.}) into dummy variables. And finally, the interaction
variables are estimated with the group of dummy variables mentioned
above. The inclusion of the interaction of the numeric variables are
validate using a linear model, and the results show that the p- values
of the parameters for the interactions are less than 0.05.

\begin{figure}
\centering
\includegraphics[width=3.30208in,height=\textheight]{includes/reg_table.png}
\caption{Linear Model: Inclusion of Interaction Variables}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{includes/corr_matrix.png}
\caption{Correlation Matrix}
\end{figure}

After the data preparation and feature engineering processes, we
construct a final data set with 33,244 observations and 421 features.
Due to the large number of interaction variables, we decided to reduce
them and use only the interaction between a group of variables
(\emph{occrecode, educcat and prestg10}). After reduction, we develop
the modeling process with 100 variables (numerical and interaction
variables). The following figure shows the overview of the group of
variables that were estimated and used in the models.

\begin{figure}
\centering
\includegraphics{includes/feature_process.png}
\caption{Feature Engineering}
\end{figure}

@@

We use five different regression techniques to predict the individuals'
wages using the final data set. We fit every model on our training data
using the \texttt{train} function from the \texttt{caret} package. To
choose the best parameter combination we use cross-validation within all
train functions.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{formula }\OtherTok{\textless{}{-}} \FunctionTok{as.formula}\NormalTok{(}\FunctionTok{paste}\NormalTok{(}\StringTok{"realrinc \textasciitilde{}"}\NormalTok{, }
                            \FunctionTok{paste}\NormalTok{(x\_cols\_dummys, }\AttributeTok{collapse =} \StringTok{" + "}\NormalTok{)))}
\NormalTok{mod\_full }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula,}
                  \AttributeTok{data =}\NormalTok{ train, }
                  \AttributeTok{method =} \StringTok{"lm"}\NormalTok{,  }
                  \AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{, }\AttributeTok{number =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Given some extreme outliers in the income distribution of our data
(which is typical for wage data) we follow the economic literature
\autocite{Ermini} and fit a logistic model to our data, using the log of
\texttt{realrinc} as our dependent variable. This results in a higher
RMSE than the linear model (25,909.82). Still, the linear model yields
predictions of \texttt{realrinc} that are negative, what is implausible.
Using the log of \texttt{realrinc} only positive predictions are
generated. Therefore, we prefer using the log for the following
predictions.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{formula2 }\OtherTok{\textless{}{-}} \FunctionTok{as.formula}\NormalTok{(}\FunctionTok{paste}\NormalTok{(}\StringTok{"log\_realrinc \textasciitilde{}"}\NormalTok{, }
                              \FunctionTok{paste}\NormalTok{(x\_cols\_dummys, }\AttributeTok{collapse =} \StringTok{" + "}\NormalTok{)))}
\NormalTok{mod\_loginc }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula2,}
                    \AttributeTok{data =}\NormalTok{ train, }
                    \AttributeTok{method =} \StringTok{"lm"}\NormalTok{,  }
                    \AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{, }\AttributeTok{number =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\[
y = RSS + \lambda \sum^p _{j=1} \beta^2_j 
\]

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ridge }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula, }\AttributeTok{data =}\NormalTok{ train,}
  \AttributeTok{method =} \StringTok{"glmnet"}\NormalTok{, }\AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{, }\AttributeTok{number =} \DecValTok{3}\NormalTok{),}
  \AttributeTok{tuneGrid =} \FunctionTok{expand.grid}\NormalTok{(}\AttributeTok{alpha =} \DecValTok{0}\NormalTok{, }\AttributeTok{lambda =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.001}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

Perform a ridge regression to predict real incomes yields an RMSE of
26,062.58.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lasso }\OtherTok{\textless{}{-}} \FunctionTok{train}\NormalTok{(formula, }\AttributeTok{data =}\NormalTok{ train, }
       \AttributeTok{method =} \StringTok{"glmnet"}\NormalTok{, }\AttributeTok{trControl =} \FunctionTok{trainControl}\NormalTok{(}\AttributeTok{method =} \StringTok{"cv"}\NormalTok{), }
       \AttributeTok{tuneGrid =} \FunctionTok{expand.grid}\NormalTok{(}\AttributeTok{alpha =} \DecValTok{1}\NormalTok{, }\AttributeTok{lambda =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.001}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

To perform the lasso regression in \texttt{R} we set \(\alpha\) equal to
one. The lasso regression yields a lower RMSE than the ridge regression
(25,918.25 vs.~26,062.58).

In a final step we use elastic net regularization which is a linear
combination of ridge and lasso regression. The elastic net
regularization allows \(\alpha\) to vary between zero and one. This
results in an RMSE of 25,918.81.

@@

\hypertarget{random-forest}{%
\subsection{Random Forest}\label{random-forest}}

According to the literature review, tree-based models perform
considerably well in range in predicting income ranges (classification
problems). For this reason, we use the random forest algorithm for
modeling. This experiment use the income without logarithm
transformation, the initial features and some interaction variables.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnr\_rdgrid\_search }\OtherTok{=} \FunctionTok{tnr}\NormalTok{(}\StringTok{"random\_search"}\NormalTok{, }\AttributeTok{batch\_size =} \DecValTok{10}\NormalTok{)}
\NormalTok{rsmp\_cv3 }\OtherTok{=} \FunctionTok{rsmp}\NormalTok{(}\StringTok{"cv"}\NormalTok{, }\AttributeTok{folds =} \DecValTok{3}\NormalTok{)}
\NormalTok{msr\_ce }\OtherTok{=} \FunctionTok{msr}\NormalTok{(}\StringTok{"regr.rmse"}\NormalTok{)}
\NormalTok{learner }\OtherTok{=} \FunctionTok{lrn}\NormalTok{(}\StringTok{"regr.lightgbm"}\NormalTok{,}
              \AttributeTok{boosting =} \StringTok{"rf"}\NormalTok{,}
              \AttributeTok{objective =} \StringTok{"regression"}\NormalTok{,}
              \AttributeTok{max\_depth =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{12}\NormalTok{,}\DecValTok{1}\NormalTok{)),}
              \AttributeTok{num\_leaves =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{270}\NormalTok{,}\DecValTok{280}\NormalTok{,}\DecValTok{1}\NormalTok{)),}
              \AttributeTok{num\_iterations  =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{132}\NormalTok{,}\DecValTok{137}\NormalTok{,}\DecValTok{1}\NormalTok{)),}
              \AttributeTok{min\_data\_in\_leaf =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{70}\NormalTok{,}\DecValTok{85}\NormalTok{,}\DecValTok{1}\NormalTok{)),}
              \AttributeTok{min\_data\_in\_bin =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{10}\NormalTok{,}\DecValTok{15}\NormalTok{, }\DecValTok{1}\NormalTok{)),}
              \AttributeTok{feature\_fraction\_bynode =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{0.4}\NormalTok{,}\FloatTok{0.1}\NormalTok{)),}
              \AttributeTok{bagging\_fraction =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\FloatTok{0.2}\NormalTok{,}\FloatTok{0.3}\NormalTok{,}\FloatTok{0.1}\NormalTok{)),}
              \AttributeTok{bagging\_freq =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{3}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{)),}
              \AttributeTok{feature\_fraction =} \FunctionTok{to\_tune}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\FloatTok{0.7}\NormalTok{,}\FloatTok{0.9}\NormalTok{,}\FloatTok{0.1}\NormalTok{)),}
              \AttributeTok{convert\_categorical =} \ConstantTok{TRUE}\NormalTok{,}
              \AttributeTok{force\_col\_wise =} \ConstantTok{TRUE}\NormalTok{,}
              \AttributeTok{verbose =} \DecValTok{1}\NormalTok{,}
              \AttributeTok{num\_threads =} \DecValTok{5}
\NormalTok{)}

\NormalTok{instance.rf }\OtherTok{=} \FunctionTok{tune}\NormalTok{(}
  \AttributeTok{tuner =}\NormalTok{ tnr\_rdgrid\_search,}
  \AttributeTok{task =}\NormalTok{ task,}
  \AttributeTok{learner =}\NormalTok{ learner,}
  \AttributeTok{resampling =}\NormalTok{ rsmp\_cv3,}
  \AttributeTok{measures =}\NormalTok{ msr\_ce,}
  \AttributeTok{term\_evals =} \DecValTok{150}\NormalTok{,}
  \AttributeTok{store\_models =} \ConstantTok{FALSE}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Due to the number of hyper parameters in random forest, we use the
\texttt{lightgbm} and \texttt{mlr3} packages to model and tune the hyper
parameter in a time-optimal way. After many attempts, we found the best
hyper parameters that produced the lowest error. The hyper parameters
tuned were \texttt{max\_depth}, \texttt{num\_leaves},
\texttt{num\_iterations} and others. In addition, we used a random
search approach in the tuning process with a cross-validation of 3. Once
we found the model along with the hyper parameters with lowest error in
the cross-validation, we proceeded to evaluate this model on the test
data set and obtained a RMSE of 31,745.54. The hyper parameters of the
best model are shown in the following table.

{[}best hyper parameter of RF{]}

In addition, we can analyze the importance of the features within the
modeling process using random forest, and these results are shown in the
following table.

{[}feature importance of RF{]}

\hypertarget{gradient-boosting}{%
\subsection{Gradient Boosting}\label{gradient-boosting}}

Gradient Boosting also has many hyper parameters like random forest and
we decided to apply a similar approach to the previous algorithm to find
the best model with the hyper parameters (random search approach in the
tuning process with a cross-validation of 3). Furthermore, the
experiments with Gradient Boosting use the target variable without
logarithm transformation.
